Using Scagnostics to find Interesting Projections of Multivariate Data

### 2018-09-27 scagnostics.Rmd
Filling in text.

### Looking down the pipe
As seen before (reference figure) MIC or TIC should be able to find the view down the pipe. The optimisation itself is however expected to be challenging because one has to be fairly close to the optimal projection to observe the increase in index value. We therefore suggest a two step procedure for the uncovering of such "small scale structures". The first step consists in "scouting" projections. Here we use the "search_better" optimisation available with the tourr package, with a standard search window $\alpha = 0.5$ and disabeling the cooling (in practice this can be done by fixing the cooling parameter to $1$), together with a large number of tries (setting max.tries to 5000 here). The resulting search will quickly steer away from the most unintresting projections, followed by a broad scan of the thus identified region. While this does not maximise the index function, it provides an informed starting plane that can be used in a second step where we use the "search_geodesic" optimisatino to improve the optimisation.
The described procedure is CPU time intensive, however it is promising that we manage to uncover such small scale structure in the distribution. On the other hand no "guarantee" can be provided for uncovering the special view. Parameters for the scouting phase will need to be adapted to different examples, and with increasing number of dimensions it may be better to consider several runs with different starting projections.
This gives some indication on how to handle difficult datasets. As we will see below most of the time standard "search_geodesic" optimisation will be efficient enough to uncover the interesting views.

@@

```{r pipeFirstRun,fig.width=6, fig.height=7, fig.cap="Guided tour optimising MIC index for Pipe 1000 dataset."}
pipeTourRes <- getProj(pipe1000, pipeTourFull, "Pipe", 1000)
pipeTourResMelt <- melt(pipeTourRes, id=c("t", "name", "size"))
ggplot(pipeTourResMelt, aes(x=t, y=value)) +
  geom_line() + 
  facet_wrap(~variable, ncol=1, scales = "free_y")
```

```{r testpipe}
iLast <- length(pipeTourFull)
fProj <- pipeTourFull[[iLast]]
dProj <- as.tibble(pipeResc %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point()
```

@@

```{r testpipe2}
iLast <- length(pipeTourFull2)
fProj <- pipeTourFull2[[iLast]]
dProj <- as.tibble(pipeResc %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point()
```

### Finding sine waves
```{r findsine, fig.width=6, fig.height=7, fig.cap="Guided tour optimising TIC index for Sine 1000 dataset.", results="hide"}
set.seed(2018)
sineResc <- rescale(sin1000) #unselect 2 variables to make optimisation easier
sineTour <- save_history(sineResc,
                         guided_tour(splineIndex()))
sineTourFull <- as.list(interpolate(sineTour))
sineTourRes <- getProj(sin1000, sineTourFull, "Sine", 1000)
sineTourResMelt <- melt(sineTourRes, id=c("t", "name", "size"))
ggplot(sineTourResMelt, aes(x=t, y=value)) +
  geom_line() + 
  facet_wrap(~variable, ncol=1, scales = "free_y")
```

```{r testsine}
iLast <- length(sineTourFull)
fProj <- sineTourFull[[iLast]]
dProj <- as.tibble(sineResc %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point()
```

>>>>

### Looking down the pipe
As seen before (reference figure) MIC or TIC should be able to find the view down the pipe. The optimisation itself is however expected to be challenging because one has to be fairly close to the optimal projection to observe the increase in index value. We therefore suggest a two step procedure for the uncovering of such "small scale structures". The first step consists in "scouting" projections. Here we use the "search_better" optimisation available with the tourr package, with a standard search window $\alpha = 0.5$ and disabeling the cooling (in practice this can be done by fixing the cooling parameter to $1$), together with a large number of tries (setting max.tries to 5000 here). The resulting search will quickly steer away from the most unintresting projections, followed by a broad scan of the thus identified region. While the lack of cooling means that this will not converge towards the maximum, each step in this first search will increase the index value, thus it can be used to provide an informed starting plane that can be used in a second step where we use the "search_geodesic" optimisatino to improve the index further towards the maximum.
The described procedure is CPU time intensive, however it is promising that we manage to uncover such small scale structure in the distribution. On the other hand no "guarantee" can be provided for uncovering the special view. Parameters for the scouting phase will need to be adapted to different examples, and with increasing number of dimensions it may be better to consider several runs with different starting projections.
This gives some indication on how to handle difficult datasets. As we will see below most of the time standard "search_geodesic" optimisation will be efficient enough to uncover the interesting views.

The results are shown below, first Fig. \ref{fig:pipeFirstRun} shows how the considered index values evolve in the scouting phase. Note that this is the interpolated path and there are small local maxima and minima along the index value, especially because we keep the step size in the guided tour large. We can see the clear maximum of the TIC index in the final projection, as well as clear minimum of the convex index that carries similar information On the other hand we see that MIC or the dcor2d index are not suitable for uncovering the special view here.

@@

```{r pipeFirstRun,fig.width=6, fig.height=7, fig.cap="Index value on tour path when scouting for high TIC index values."}
pipeTourRes <- getProj(pipe1000, pipeTourFull, "Pipe", 1000)
pipeTourResMelt <- melt(pipeTourRes, id=c("t", "name", "size"))
ggplot(pipeTourResMelt, aes(x=t, y=value)) +
  geom_line() + 
  facet_wrap(~variable, ncol=1, scales = "free_y")
```

We next show in Fig \ref{fig:testpipe} the final view corresponding to the highest index found in the TIC scouting phase. While this is clearly not the optimal view of the pipe feature, we can already guess the type of dependence in the distribution, and we expect to be able to find the maximum when using this plane as a starting point of the geodesic optimisation. Indeed performing a standard guided tour with the TIC index and the identified starting plane efficiently identifies the optimal view shown in Fig ...
((Probably should remove second trace plot as it does not give much information, maybe show final view in same figure as final scouting view...))

```{r testpipe, out.width=".5\\textwidth", fig.cap="Final view discovered by the scouting phase."}
iLast <- length(pipeTourFull)
fProj <- pipeTourFull[[iLast]]
dProj <- as.tibble(pipeResc %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point()
```

@@

```{r testpipe2, out.width=".5\\textwidth"}
iLast <- length(pipeTourFull2)
fProj <- pipeTourFull2[[iLast]]
dProj <- as.tibble(pipeResc %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point()
```

### Finding sine waves
We next work with distribution 3 and aim to identify the view showing functional dependence using the splines2d index. Given the results from Fig 6 ((add proper ref)), we expect to be able to efficiently identify this view without relying on the two step procedure described above. Indeed the guided tour quickly converges in this case, the evolution of the index functions is shown in Fig \ref{fig:findsine}. Here there are four index functions that appear to be suitable for the optimisation: splines2d, dcor2d, MIC and TIC. On the other hand, while some of the scagnostics indices are reaching their maximum/minimum in the final projection, they do not show a smooth increase/decrease thus preventing efficient optimisation when considering them as index functions. The final view identified by the guided tour is shown in Fig \ref{}

```{r findsine, fig.width=6, fig.height=7, fig.cap="Guided tour optimising the splines2d index for the Sine 1000 dataset.", results="hide"}
set.seed(2018)
sineResc <- rescale(sin1000) #unselect 2 variables to make optimisation easier
sineTour <- save_history(sineResc,
                         guided_tour(splineIndex()))
sineTourFull <- as.list(interpolate(sineTour))
sineTourRes <- getProj(sin1000, sineTourFull, "Sine", 1000)
sineTourResMelt <- melt(sineTourRes, id=c("t", "name", "size"))
ggplot(sineTourResMelt, aes(x=t, y=value)) +
  geom_line() + 
  facet_wrap(~variable, ncol=1, scales = "free_y")
```

```{r testsine, out.width=".5\\textwidth", fig.cap="Final view identified by the optimisation."}
iLast <- length(sineTourFull)
fProj <- sineTourFull[[iLast]]
dProj <- as.tibble(sineResc %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point()
```
